---
title: "Relationship Between Turnout in Federal Elections and the Decline in Local Journalism"
author: 
  - Jet McCullough
  - William Chan
thanks: "Code and data are available at: [https://github.com/jethroid/INF2167-Proposal-Group-2](https://github.com/jethroid/INF2167-Proposal-Group-2)."
date: August 1, 2026
date-format: long
format: pdf
number-sections: true
bibliography: paper_references.bib
---

# Abstract
Local news is important to the health of democracies because it keeps voters informed and holds politicians accountable. Previous research has shown that the decline of local news reduces voters’ knowledge of local candidates and affairs, which can discourage civic engagement in communities. The closure of local news outlets across Canada therefore carries potentially significant political implications, one of which concerns the political participation of the population. To explore and understand the relationship between local journalism and political participation in Canada, we curated a dataset documenting changes in local news outlets and voter turnout across three federal elections and performed a linear regression analysis. The results found no significant relationship between changes in local news outlets and voter turnout. However, this finding by no means suggests that the decline of local news in Canada has no effect on political participation or political culture. Future research can test alternative measures of political participation and benefit from the dataset developed in this study.  


# Introduction

Local journalism in Canada is experiencing a sharp decline, with 2.5 million Canadians, or 7 percent of the population, living in postal codes with one or no local media outlets [@macdonald_news_2025]. As local news provides important information about community affairs to electorates, the continued loss of local news can leave citizens less informed about local issues, potentially discouraging political participation and thereby threatening the democratic system. Despite these crucial political implications, existing studies on local news deserts have overwhelmingly focused on the United States, which has a different electoral system and political culture from Canada. Therefore, our project aims to contribute to the understanding of the relationship between local news and political participation among Canadians by answering the following research question:

**How does the net trajectory of the local news market affect a riding’s voter turnout relative to the national baseline?**

Our hypothesis is that ridings that experience more negative changes in their local news market will have lower voter turnout than the national average. To test this hypothesis, we use federal election results from Elections Canada and local news outlet data from the Local News Research Project, led by researchers at Toronto Metropolitan University [@lindgren_local_nodate] to conduct a linear regression analysis quantifying the relationship between changes in the local news and voter turnout in Canada. The project outlined below includes the creation of this mixed dataset and  regression model, which form useful first steps to exploring the relationship between local news and electoral results in Canada.

# Relevant Literature

The loss of local news in Canada affects a sizable portion of the population, but its effects are not distributed equally. @barber_left_2026 found that news outlet closures are more likely to occur in “left-behind” places, places that are “small, shrinking, peripheral", reflecting the broader inequalities driven by globalization. Economic decline in these places has resulted in unemployment, business closures, and retrenchment in public services, all of which have likely undermined the vitality of local news outlets (p. 11). @barber_left_2026 also highlighted the political consequences of news closures in left-behind places, arguing that voter turnout is likely to decline because of a lack of access to information and pessimism that their votes do not matter in elections. Additionally, the loss of local news can widen the space for misinformation and conspiracy theories, contributing to greater support for populist and polarizing agendas (p. 12).

@hayes_decline_2018 argue that local news helps electorates develop the political knowledge necessary to participate in the democratic process, such as voting. Using U.S. House campaigns in 2010 and 2014, they showed that citizen engagement declined as coverage of House races declined. With less reporting on candidates and their campaigns, fewer voters were able to rate the candidates in a survey (p. 335). The authors argue that the deterioration of political knowledge is concerning, as an informed electorate that is able to hold elected officials accountable improves the representation and legitimacy of the political system.

Besides being a source of information, local journalism can serve as an active venue for community members to participate in the political process. @magasic_examining_2023 provide examples from Australian towns where residents expressed their opinions and advocated for social issues, such as cuts to welfare services. Local journalists help maintain the relationship between the government and the public, allowing for the exchange of information through sections such as “Letters to the Editor.” Local news fosters a sense of community and thereby promotes civic engagement.

@haddock_local_2026 expanded on this idea by arguing that the loss of local news negatively affects community social capital, resulting in a feedback loop that weakens trust between voters and institutions. Reduced coverage of local issues and elected officials affects the community’s ability to monitor crime and intensifies polarization in rural communities, as residents turn to national, partisan media for political information.

# Dataset and Methods

We rely on datasets from three sources. The local news dataset was downloaded from the Local News Research Project, which releases its local news map report as .xlsx files on a bi-monthly basis, tracking changes in local news outlets in Canada since 2008 [@lindgren_local_nodate]. For our analysis, we obtained the data as of June 8, 2026, which contains 1,337 news outlets and 18 variables. It is a spatial dataset because it includes the coordinates of individual outlets, which are crucial for our subsequent step of merging the data with the election results. Transition Type is another key variable that records changes to news outlets and groups them into eight categories: “closed,” “shifted to online,” “closed due to merger,” “decrease in service,” “new,” “increase in service,” “new outlet produced by merger,” and “daily becomes a community paper.” The Date of Change column is used to map each change to the corresponding election period. These categories are re-coded into two categories `positive` and `negative` for measuring the changes of local news.

The second dataset we use consists of Canadian federal election results for 2015, 2019, and 2021, downloaded from Elections Canada [@electionscanada]. The agency publishes election results by polling station for each election year as CSV files stored in a compressed folder. Total Votes and Electors are the key variables used to calculate voter turnout. Each CSV file also includes the Electoral District Number, which identifies the riding and is used to merge the election results with the local news dataset.

The Federal Electoral District Boundary Files are the third dataset we use to merge the local news and election datasets [@fed_boundaries]. We downloaded the boundary file for the 2021 census year, which applies to all three elections in our study. The file is in shapefile (.shp) format and contains geometric attributes for polygons delineating the boundaries of federal electoral districts. In addition to spatial attributes, the file includes metadata for each riding, particularly the FEDDGUID_CEFIDUGD column, which refers to the Dissemination Geography Unique Identifier for electoral districts [@fed_boundaries_dict]. The last five digits of this column correspond to the Electoral District Number in the election data and are therefore used for merging.

To merge the local news and election datasets, we first convert the local news table into a spatial object using the Simple Features (sf) package [@simple_features]. The local new data is assigned the WGS84 projection to ensure correct alignment on the map. We then merge the news data with the boundary file by conducting a spatial join using the "within" relationship. Each observation in the news dataset is assigned an electoral district ID from the shapefile if it falls within the corresponding polygon. With the electoral district ID attached, the news dataset can then be merged with the election results using a standard left join in the tidyverse package.

To quantify the relationship between voter turnout and changes in local news outlets, we ran the following linear regression model:

$$turnout\_diff_{it} = \beta_0 + \beta_1 \cdot cumulative\_news_{it} + \epsilon_{it}$$

The outcome variable, $turnout\_diff_{it}$, is the difference between the turnout of a specific riding $i$ in the election year $t$ and the overall national turnout in that same election year. The predictor, $cumulative\_news_{it}$, represents the net running sum of local news changes for a specific riding $i$ up to the date of the election $t$. This sum is calculated by assigning an integer value corresponding to the cumulative changes of news in a riding from 2008 to the date of the election in question, where a negative news change ("closed," "shifted to online," "closed due to merger," "decrease in service") counts as -1, a positive change ("new," "increase in service") counts as +1, and relatively neutral changes ("new outlet produced by merger," "daily becomes a community paper") do not affect the total.

# Results and Findings

```{r}
#| echo: false
#| include: false

library(tidyverse)
library(tinytable)
library(here)
library(scales)
library(modelsummary)
library(hrbrthemes)

withdupes <- read_csv(here("data", "02-analysis_data", "news_election_before_dupe_removal.csv"))
newselec <- read_csv(here("data", "02-analysis_data", "news_election.csv"))

```

## Regression

Our attempts at fitting a regression model to the data hardly produced overwhelming evidence of a correlation between negative changes to local news outlets and reduced turnout. Our first model, following the formula outlined above, returned a coefficient of -0.001 and an R-Squared value of 0.002. Other models were tried, with variations on the news change and turnout variables:

2) A model that uses changes to news and difference of relative turnout election-by-election (2015-19, 2019-21) rather than cumulative news changes to date and turnout difference in that election year:

$$\Delta turnout\_diff_{it} = \beta_0 + \beta_1 \cdot \Delta news_{it} + \epsilon_{it}$$
Where $\Delta turnout\_diff_{it}$ is the difference from the previous election year in the value of riding turnout minus national average turnout, and $\Delta news_{it}$ is the change in news from the previous election year, i.e. the cumulative news score minus the previous year's score.


3) A model similar to the first attempted one, with cumulative news changes and turnout difference, but where the score of news changes excludes online news sources, and includes only old media. Media type is an element of the local news dataset that we thought may have gone uncaptured by our model. Since the final form of our merged dataset contains observations that are neatly per-riding and per-year, and the media type variable relies on observations that are per-news-change, we attempted to account for media type by changing how the cumulative news integer value is calculated rather than attempting to use it as another predictor in the model.

4) A model similar to 2), looking at election-by-election changes, but where only old media are included as in 3).

None of these models showed much evidence that the news change and turnout variables respectively used were closely correlated, however. The model described in 2) produced the most pronounced slope (-0.003) and the highest R-Squared value (0.017), and is pictured below (Figure 5.1.1). As is apparent, these results cannot be reasonably interpreted as overwhelming evidence of a connection between local news and political outcomes on a national level.

```{r}
#| echo: false
#| warning: false

ggplot(data = newselec, 
       aes(x = change_last, y = turnout_diff_last)) +
  geom_point(alpha = 0.3, color = "darkgray") +
  
  #overlay the regression line with a 95% confidence interval
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  labs(
    x = "News Change From Previous Election",
    y = "Turnout Difference",
    caption = "Figure 5.1.1. News changes from previous election as a predictor of turnout changes."
  ) +
  
  theme_minimal()
```


## Merged Dataset

The methods outlined above resulted in a reproducible workflow that results in a dataset that has local news outlet information and elections information in one place, along with the variables added by us. The code for merging datasets based on spatial join is hosted on our Github repository and can be adapted or expanded to include future and past election and news information, or other data that relies on census or geographic divisions. We can now more easily explore the data for the elections and news outlets we have already merged, and use our new variables beyond modeling a regression.  We can use the positive and negative values we assigned to types of news changes, for example, to explore what types of media have seen the greatest number of negative changes, as in Table 1. As we can see, a majority of negative changes were to community papers.

```{r}
#| echo: false

withdupes |>
  group_by(media_type) |>
  summarize(neg_number = -sum(type_val[type_val == -1])/3) |>
  arrange(desc(neg_number)) |>
  tt(caption = "Media Types With The Most Negative News Changes", colnames = FALSE, width = 1) |>
  style_tt(i = "caption", bold = TRUE) |>
  style_tt(i=1, j=2, background = "#FFADAD") |>
  style_tt(i=6:9, j=2, background = "#FFE0E0") |>
  style_tt(i=2:5, j=2, background = "#FFC2C2")
```

Having the elections and news data in one dataset, we can examine the relationships among these data in different ways. Each riding/election year now has a score which is a measurement of news changes that happened within its boundaries between 2008 and the year in question. Tables 2 and 3 below give us a look at all ridings that have a cumulative score of -5 or lower as of 2021, i.e. those whose local news has been most negatively affected on the whole, alongside an equivalent length list of ridings with the greatest negative difference in turnout from the national average. This comparison would also not seem to provide evidence of a strong connection, though three ridings are in common among these lists, as are highlighted below.

Since we have the data on local news at our fingertips, we have the ability to look at the stories these ridings point to. Between the 2019 and 2021 election, Coast of Bays--Central--Notre Dame, for example, saw the closure of "The Central Voice" due to pressures resulting from the pandemic. As the descriptions included in the news data inform us, "The Central Voice" was a free paper that had served central Newfoundland and had itself been the result of a merger of four community papers in the area. Though the riding had relatively low turnout in all three elections for which we have data, the period between the 2019 and 2021 elections did see some sort of shift in the political culture in this riding, as the result shifted from a solid Liberal victory in 2019 to a narrow win for the Conservatives in 2021.

```{r}
#| layout-ncol: 2
#| echo: false

# make table of worst news changes as of 2021

newselec |>
  filter(election_year == 2021) |>
  select(fedename, cumulative_news) |>
  arrange(cumulative_news) |>
  slice_head(n=23) |>
  tt(caption = "Ridings With The Most Negative News Impact Score", colnames = FALSE, width = 1) |>
  style_tt(i = "caption", bold = TRUE) |>
  style_tt(i=1:5, j=2, background = "#F5C7FF") |>
  style_tt(i=13:23, j=2, background = "#FCEBFF") |>
  style_tt(i=6:12, j=2, background = "#F9DBFF")|>
  style_tt(i=12, j=1, background = "#D1DAFF") |>
  style_tt(i=15, j=1, background = "#D1DAFF") |>
  style_tt(i=20, j=1, background = "#D1DAFF")

# make table of lowest turnout relative to ntl. avg.

newselec |>
  filter(election_year == 2021) |>
  select(fedename, turnout_diff) |>
  arrange(turnout_diff) |>
  slice_head(n=23) |>
  tt(caption = "Ridings With The Lowest Turnout Relative to National Average", colnames = FALSE, width = 1) |>
  style_tt(i = "caption", bold = TRUE) |>
  style_tt(i=1:5, j=2, background = "#FFDB9E") |>
  style_tt(i=13:23, j=2, background = "#FFF6E5") |>
  style_tt(i=6:12, j=2, background = "#FFEBC7") |>
  style_tt(i=23, j=1, background = "#D1DAFF") |>
  style_tt(i=15, j=1, background = "#D1DAFF") |>
  style_tt(i=19, j=1, background = "#D1DAFF")
```

Another area of news changes we have not touched on yet are the ones that we have deemed positive, i.e. new outlets or expansions of existing outlets. With our merged dataset, we can examine the ridings that have had the most positive change overall. As we can see from Figure 5.2.1 below, the highest positive cumulative news score in 2021 was in the riding of Québec. Looking again at our description variable from the Local News Research Project, we see in fact that the high score is a result of multiple "hyperlocal" neighbourhood-oriented web outlets being formed in this period, and a then-new Québec City TV newscast from Noovo TV network. Turnout in this riding stayed in the high sixties and slightly above the national average in all three years we have data for.

```{r}
#| echo: false

colours <- c("#008A45", "#CCA500", "#A80084", "#295BFF", "#A30500")

newselec |>
  filter(election_year == 2021) |>
  select(fedename, cumulative_news) |>
  arrange(desc(cumulative_news)) |>
  slice_head(n=5) |>
  ggplot(aes(x = reorder(fedename, -cumulative_news), y = cumulative_news, fill = fedename)) +
  geom_col(show.legend = FALSE) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, size = 7, face = "bold")) +
  scale_fill_manual(values = colours) +
  scale_x_discrete(labels = c("London North Centre" = "London \n North Centre",
                              "Esquimalt--Saanich--Sooke" = "Esquimalt\n --Saanich--\n Sooke",
                              "Scarborough--Guildwood" = "Scarborough\n --Guildwood")) +
  labs(
    x = NULL,
    y = "News Score",
    caption = "Figure 5.2.1. Ridings with the highest news scores."
  )
```

# Discussion and Conclusion

## Discussion

Needless to say, that this project did not produce strong evidence of a link between local news decline and turnout in federal elections is not grounds to conclude that the alarming amounts of local media closures in recent decades does not have an effect on the political culture in Canada or is nothing to worry about. Because of time constraints, our project has been limited in scope to our one research question, but turnout is not the only measure of political engagement or political health. Next steps for a project such as this could include the addition of more data and more variables to measure things like polarization and level of discourse. Because of the complexity of working with spatial data across redistricting periods, our merged dataset only includes results from three general elections working with the same electoral boundaries, and does not include data from provincial or municipal elections. We represent here only a fraction of the political landscape of Canada over this period, but our code and dataset serves as a model or template for further efforts to connect Canadian electoral and news data in the future.

Due to similar limitations, our plan to merge additional census data into our dataset was put to the side. As a result, our regression is limited to local news changes as the sole predictor rather than exploring predictors such as income and education levels as would have been the course had time permitted. Whether factors like these or others are confounding variables, obscuring a correlation that is there, or whether there really is little relationship as our model suggests, remains a question worth exploring. The regression may also be improved by a more sophisticated measurement of news change than our simple integer values: our value does not account for differences in population among ridings, for instance, which may inflate the negative numbers of urban ridings, though viewing Tables 2 and 3 above suggests that urban and rural ridings are both represented in the categories of high local news closures and low turnout.

Building on our project with more data and better models can illuminate this question of confounding variables but also disentangle other factors from turnout--might we see, for example, contrary to expectations, reduced trust in democracy without depressed turnout at the federal elections level? Next steps could include studying differences among political engagement with different levels of government: as mentioned in Section 2, a reduction in local news can increase dependence on national, partisan media, and meanwhile we could even see reduced turnout in municipal elections as predicted by @barber_left_2026 (p. 11).

## Conclusion

The project outlined in this paper did not produce a regression model that convincingly explains the relationship between political engagement and changes to local news in Canada. The dataset produced in the course of the project offers promise for exploring this relationship, however, and points to the ridings where these factors may intersect, which have their own individual political stories that are worth exploring. The project forms a starting point for future explorations in the Canadian political and news media space. The code to create this dataset, located in R Script files on our Github repository, can be executed independently, and added to and repurposed to add geographical data for other configurations of political boundaries besides the 2013 map, census data with potentially explanatory variables such as income and education, electoral results for provincial and municipal elections, and future local news changes. Future researchers will be able to expand the pool of connected data to draw from by similar means and explore the relationship between local news and political engagement in a more sophisticated and perhaps conclusive way.

\newpage

# References
