---
title: "The Demise of Local Journalism and Democratic Health in Canada"
author: 
  - Jet McCullough
  - William Chan
thanks: "Code and data are available at: [https://github.com/jethroid/INF2167-Proposal-Group-2](https://github.com/jethroid/INF2167-Proposal-Group-2)."
date: June 18, 2026
date-format: long
format: pdf
number-sections: true
bibliography: proposal_references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(here)
```

# Context & Literature Review

Local journalism has been consistently declining in Canada. A 2025 report found that 2.5 million Canadians live in areas with almost no local news coverage, and the remaining outlets continue to face the threat of closure [@macdonald_local_2025]. @lindgren_what_2019 argues that local news acts as a "civic watchdog" that helps residents evaluate politicians in their ridings by scrutinizing the records of local candidates and incumbents. The demise of local journalism is thus detrimental to the health of democracy in Canada.

The crisis of local news deprivation is not unique to Canada. @haddock_local_2026 found a similar trend in the United States, where the number of counties without local news doubled to more than 200 between 2010 and 2016. This decline is associated with increases in non-violent crime, lower voter turnout, and greater political polarization [@haddock_local_2026], as residents rely more heavily on national media, which tends to be more partisan and less focused on local issues, thereby amplifying ideological divisions and political apathy toward local politics.

Likewise, @moskowitz_local_2021 argues that elections have become more nationalized in the United States as voters increasingly rely on national, partisan media to assess the performance of politicians. Greater local news coverage leads to higher levels of voter knowledge about local incumbents and candidates, allowing voters to make assessments that differ from their "national judgement on the presidential race" in the form of split-ticket voting (p. 126). @moskowitz_local_2021 posits that highly nationalized elections undermine the accountability of officeholders, as their performance is more likely to be judged by the popularity of their party and president rather than by their competence and service delivery (p. 127).

Echoing @moskowitz_local_2021, @ellger_local_2024 provides evidence from Germany, where they found that the disappearance of local news is correlated with an increase in the vote share of small parties. In the German party system, small parties tend to be further from the ideological center; thus, an increase in their vote shares indicates greater polarization among the electorate (p. 1266).

# Research Questions

Scholarly studies on news deprivation and its political implications mostly focus on the U.S. context, thereby revealing a research gap in the Canadian context. Drawing from existing literature, we propose two research questions:

**RQ1. Do Canadian ridings with lower levels of local news coverage exhibit lower voter turnout in federal and municipal elections than ridings with higher levels of local news coverage?**

H1: Ridings that experience a decline in local news coverage will exhibit lower voter turnout in federal and municipal elections, as reduced access to local political information decreases voter knowledge and engagement.

**RQ2. Is the decline of local news in Canada associated with higher levels of political polarization at the riding level?**

H2: Ridings that experience a decline in local news coverage will exhibit higher levels of political polarization, as voters increasingly rely on national partisan media and evaluate politics through national rather than local perspectives.

# Dataset and Methods

We will be relying on datasets from two sources. One is from the @localnews, who every two months reports changes to local media outlets, from 2008 onwards. Each row in the dataset is a change to one local news source, i.e. a paper or other media outlet, and what happened to it. There are three key variables in this dataset for the purposes of answering our research questions. One is "Transition Type," as the project maps positive changes (new papers, increases in service) as well as negative (closures, etc.). The other two key variables are "Date of Change" and "Community" (i.e. city, town, etc. of publication), which will allow us to compare these changes to election results by year and region.

The other datasets we will be using are Canadian general election results over the same period (2008-2025), downloaded from @electionscanada. These are published in two formats, one for each riding. In Format 1, each row is a poll with a unique ID number, and variables recorded include candidate/total vote amounts and number of electors. In Format 2, each row is a poll result for one candidate, though poll numbers and candidate names/parties are tidily kept as distinct variables. Because Format 2 includes each candidate's political party and vote count as separate variables, it will allow us to study polarization by calculating vote shares. Both formats include electors as a variable, a necessary number for calculating turnout, but Format 1 also includes "Total Votes" for each poll, making turnout calculation easier.

Our plan is to first correspond the election ridings to the "communities" identified in the news map data. One way we might do this is using @statcan census data: Census Profile datasets include federal electoral district IDs, as do our election results, and the Census Subdivision Names that these districts are grouped under can be matched to our communities of interest. This shortcut may not work for the 2025 results, however, as the district IDs will be affected by redistricting which occurred after the 2021 census. 
We will then calculate three new variables for each community and period between elections: 1) the positive/negative change in local news for each community, i.e. a ratio between population and news changes, 2) the change in voter turnout between elections, and 3) the level of polarization, measured using margin of victory. We would then measure statistical correlation among these continuous numeric variables across time using R. If we find statistical significance, we can calculate and compare levels/directions of correlation as the following coefficients: 

$$\Delta Turnout = \beta_0 + \beta_1\Delta News + \epsilon \\$$
$$\Delta MOV = \beta_0 + \beta_1\Delta News + \epsilon\\$$
$$\beta_0: \text{intercept;}\ \beta_1: \text{correlation coefficient}$$

# Exploratory Data Analysis

```{r}
#| include: false

news <- read_csv(here("data", "01-raw_data", "proposal_data",  "news_apr_2026.csv")) #opening our datasets
election2025 <- read_csv(here("data", "01-raw_data", "proposal_data",  "2025polls", "pollresults_resultatsbureau24070.csv")) #Saint-Maurice--Champlain riding
format2025 <- read_csv(here("data", "01-raw_data", "proposal_data",  "2025pollsformat1", "pollbypoll_bureauparbureau24070.csv")) #alt. format
```

We will explore our data using R by @citeR, and the packages here, by @citehere and tidyverse, by @citetidyverse. Our local news dataset tells us that the majority of changes tracked in Canadian local news are negative, i.e. not new outlets being created.

```{r}
#| echo: false

news |>
  group_by(`Transition Type`) |> #counting each type of closure, opening, shifting in service, etc.
  count()|>
  ungroup() |>
  rename(Total = n) |>
  arrange(desc(Total))
```

We can use our dataset to look at negative changes in local news over time for a particular community. The following line plot shows that 2015 and 2019 saw spikes in local news reduction in the city of Toronto. 

```{r}
#| echo: false

torontonews <- news |> #new dataset: negative changes in Toronto
  filter(Community == "Toronto, Ontario") |>
  filter(`Transition Type` != "new" & `Transition Type` != "increase in service") 
torontonews |> 
  mutate(
    date = year(as.Date(torontonews$`Date of Change`))) |> #simplifying to year-by-year
  group_by(date) |> #getting no. of observations per year
  count() |>
  ungroup() |>
  ggplot(aes(x = date, y = n)) + 
  geom_line(colour = "navy") + #lineplot, choosing colour
  scale_x_continuous(breaks = scales::pretty_breaks(n=9)) +
  labs(x = "Year", y = "Number of Negative Changes", 
       title = "Local News Sources Lost or Reduced in Toronto")
  
```

With our elections datasets, we can find the vote shares within a riding, which we will need to measure polarization. The table below shows small parties' results for our sample riding. We can also calculate turnout numbers, and compare them across polls. The histogram below shows the distribution of turnout across polls in a constituency. This visual shows us that turnout numbers between 30 and 50 percent are most typical in this case. It also shows some turnout numbers above 100 percent: filtering the dataset to see what these are shows that it is mostly mobile polls that show vote totals greater than registered electors. That calculating turnout numbers may become complicated by the inclusion of mobile polls or other unforeseen aspects of how these data are recorded is useful information to have for when we calculate turnout by riding for this project.

```{r}
#| echo: false

election2025 |> #vote totals for parties
  group_by(`Political Affiliation Name_English/Appartenance politique_Anglais`) |>
  summarize(vote_total = sum(`Candidate Vote Count/Votes du candidat`))|>
  ungroup() |>
  mutate(vote_share = vote_total/sum(vote_total)) |>
  rename(Party = `Political Affiliation Name_English/Appartenance politique_Anglais`) |>
  filter(vote_share < 0.015) |> #only small parties, 1.5% chosen arbitrarily
  arrange(vote_total) #arranging by least votes to most
  

```

```{r}
#| echo: false
#| warning: false #remove warning that some data falls outside plot or is null

format2025 |>
  mutate(turnout = `Total Votes/Total des votes`/`Electors/Électeurs`) |> #calculating turnout
  ggplot(aes(x = turnout)) +
  geom_histogram(bins = 60, fill = "purple") + #histogram of turnout levels
  scale_x_log10() + #scale for readability
  labs(x = "Turnout", 
       y = "Number of Polls", 
       title = "Distribution of Turnout in Saint-Maurice--Champlain Riding")
```

```{r}
#| include: false

format2025 |>
  filter(`Total Votes/Total des votes` > `Electors/Électeurs`)|> #what are these over 100% turnout polls?
  select(c(`Polling Division Name/Nom de section de vote`, `Total Votes/Total des votes`, `Electors/Électeurs`))
```

Working with all the riding datasets together as we will be doing for the final project, we will be able to similarly calculate and compare turnout, as well as polarization, among ridings, and then among communities, over time.

\newpage

# References
