---
title: "The Demise of Local Journalism and Democratic Health in Canada"
author: 
  - Jet McCullough
  - William Chan
thanks: "Code and data are available at: [https://github.com/jethroid/INF2167-Proposal-Group-2](https://github.com/jethroid/INF2167-Proposal-Group-2)."
date: June 18, 2026
date-format: long
format: pdf
number-sections: true
bibliography: proposal_references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(here)
```

# Context & Literature Review

Local journalism has been consistently declining in Canada. A 2025 report found that 2.5 million Canadians live in areas with almost no local news coverage, and the remaining outlets continue to face the threat of closure [@macdonald_local_2025]. @lindgren_what_2019 argues that local news acts as a "civic watchdog" that helps residents evaluate politicians in their ridings by scrutinizing the records of local candidates and incumbents. The demise of local journalism is thus detrimental to the health of democracy in Canada.

The crisis of local news deprivation is not unique to Canada. @haddock_local_2026 found a similar trend in the United States, where the number of counties without local news doubled to more than 200 between 2010 and 2016. This decline is associated with increases in non-violent crime, lower voter turnout, and greater political polarization [@haddock_local_2026], as residents rely more heavily on national media, which tends to be more partisan and less focused on local issues, thereby amplifying ideological divisions and political apathy toward local politics.

Likewise, @moskowitz_local_2021 argues that elections have become more nationalized in the United States as voters increasingly rely on national, partisan media to assess the performance of politicians. Greater local news coverage leads to higher levels of voter knowledge about local incumbents and candidates, allowing voters to make assessments that differ from their "national judgement on the presidential race" in the form of split-ticket voting (p. 126). @moskowitz_local_2021 posits that highly nationalized elections undermine the accountability of officeholders, as their performance is more likely to be judged by the popularity of their party and president rather than by their competence and service delivery (p. 127).

Echoing @moskowitz_local_2021, @ellger_local_2024 provides evidence from Germany, where they found that the disappearance of local news is correlated with an increase in the vote share of small parties. In the German party system, small parties tend to be further from the ideological center; thus, an increase in their vote shares indicates greater polarization among the electorate (p. 1266).

# Research Questions

Scholarly studies on news deprivation and its political implications mostly focus on the U.S. context, thereby revealing a research gap in the Canadian context. Drawing from existing literature, we propose two research questions:

**RQ1. Do Canadian ridings with lower levels of local news coverage exhibit lower voter turnout in federal and municipal elections than ridings with higher levels of local news coverage?**

H1: Ridings that experience a decline in local news coverage will exhibit lower voter turnout in federal and municipal elections, as reduced access to local political information decreases voter knowledge and engagement.

**RQ2. Is the decline of local news in Canada associated with higher levels of political polarization at the riding level?**

H2: Ridings that experience a decline in local news coverage will exhibit higher levels of political polarization, as voters increasingly rely on national partisan media and evaluate politics through national rather than local perspectives.

# Dataset and Methods

We will be relying on datasets from two sources. One is from the @localnews. As part of this project, raw data is published every two months with updates on changes to local media outlets, each report ranging from the present day back to 2008. There are three key variables in this dataset for the purposes of answering our research questions. One is "transition type," as the project maps positive changes (new papers, increases in service) as well as negative. The other two key variables are "date of change" and "community" (i.e. city, town, etc.), which will allow us to compare these changes to election results over time and by region.

The other datasets we will be using are Canadian general election results over the same period (2008-2025), downloaded from @electionscanada. We will be using the official voting results results in what Elections Canada calls "Format 2," or "pollresults_resultatsbureau," which includes each candidate's political party, "Political Affiliation Name_English," and vote counts, "Candidate VoteCount." These variables together allow us to measure the relative prominence of small parties, which we take to be an indicator of left-right polarization. This format also includes the number of electors for each poll, which we will need to calculate turnout, as the ratio of total votes cast to number of electors. Though these poll-by-poll data have the disadvantage of being more granular than our local news data, we choose this format because of the flexibility of having variables separate from one another, such as candidate party and names, and because it includes the number of electors in the same place as the rest of this information. None of the summary tables we looked at from Elections Canada seemed to meet these requirements.

Our plan is to first correspond the election ridings to the "communities" identified in the news map. We will then explore methods of comparing local news health and political results, especially in communities most affected by reduction in local news sources. One way is by mathematical correlation: this could perehaps be done by calculating a ratio between a community's population and an integer representing negative/positive changes over the period between general elections, then calculating both the turnout and vote share of small parties in each election, by community. Then we could measure statistical correlation among these three new continuous numeric variables for each election year suing R, and chart and compare their movement over time. Alternatively, we can reduce the scope by zeroing in on some particular communities and attempting to tell their local news stories, drawing parallels to interesting political results, seeing if these data match our hypotheses or not.

# Exploratory Data Analysis

The local news dataset can show (need to cite here package)

```{r}
news <- read_csv(here("data", "01-raw_data", "news_apr_2026.csv"))
head(news)
```


\newpage

# References
